---
title: "Reproducible Research: Peer Assessment 1"

```{r, eval=FALSE}
## First, we need to install the required packages
#install.packages("dplyr")
install.packages("Hmisc")


library(scales)
library(Hmisc)
#library(dplyr)

```

```{r setup, include = FALSE}
library(dplyr)
library(plyr)
library(ggplot2)
library(lattice)
```


## Loading and preprocessing the data
##### 1. Load the data (i.e. read.csv())
```{r, results='markup', warning=TRUE, message=TRUE}
if(!file.exists('activity.csv')){
    unzip('activity.zip')
}
activityData <- read.csv('activity.csv')
```

##### 2. Process/transform the data (if necessary) into a format suitable for your analysis
```{r}
activityDataCleaned <- mutate(activityData, dateCleaned = as.POSIXct(activityData$date, format = "%Y-%m-%d"))
```


```{r echo = TRUE, message=FALSE, warning=FALSE}
# Addind a new field - day
activityDataCleaned$day <- weekdays(as.Date(activityDataCleaned$date))
head(activityDataCleaned)
```

## What is mean total number of steps taken per day?
### Calculate the total number of steps taken per day
```{r}
### summarizing total steps per date
sumTable <- aggregate(activityDataCleaned$steps ~ activityDataCleaned$dateCleaned, FUN=sum, )
colnames(sumTable)<- c("Date", "Steps")
```

##### 1. Make a histogram of the total number of steps taken each day
```{r}
hist(sumTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day")
```

##### 2. Calculate and report the mean and median total number of steps taken per day
```{r}
stepsByDayMean <- as.integer(mean(sumTable$Steps))
stepsByDayMedian <- as.integer(median(sumTable$Steps))
```
* Mean: `r stepsByDayMean`
* Median:  `r stepsByDayMedian`

## What is the average daily activity pattern?
### Make a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
##pulling data without nas
clean <- activityDataCleaned[!is.na(activityDataCleaned$steps),]

##create average number of steps per interval
intervalTable <- ddply(clean, .(interval), summarize, Avg = mean(steps))
head(intervalTable)
```
##### 1. Make a time series plot
```{r}
##Create line plot of average number of steps per interval
p <- ggplot(intervalTable, aes(x=interval, y=Avg), xlab = "Interval", ylab="Average Number of Steps")
p + geom_line()+xlab("Interval")+ylab("Average Number of Steps")+ggtitle("Average Number of Steps per Interval")
```
##### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
##Maximum steps by interval
maxSteps <- max(intervalTable$Avg)
##Which interval contains the maximum average number of steps
timeMostSteps <- intervalTable[intervalTable$Avg==maxSteps,1]
maxSteps
```

* The maximum number of steps for a 5-minute interval: `r round(maxSteps)`
* Most Steps at: `r timeMostSteps` interval

## Imputing missing values

##### 1. Calculate and report the total number of missing values in the dataset 
```{r}
##Number of NAs in original data set
numMissingValues <- nrow(activityDataCleaned[is.na(activityDataCleaned$steps),])
```

* Number of missing values: `r numMissingValues`

##### 2. Devise a strategy for filling in all of the missing values in the dataset.
##### My strategy for filling in NAs will be to substitute the missing steps with the average 5-minute interval based on the day of the week.

```{r}
## Create the average number of steps per weekday and interval
avgTable <- ddply(clean, .(interval, day), summarize, Avg = mean(steps))
```

```{r}
## Create dataset with all NAs for substitution
nadata<- activityDataCleaned[is.na(activityDataCleaned$steps),]
## Merge NA data with average weekday interval for substitution
newdata<-merge(nadata, avgTable, by=c("interval", "day"))
```
##### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
## Reorder the new substituded data in the same format as clean data set
newdata2<- newdata[,c(6,4,1,5,2)]

colnames(newdata2)<- c("steps", "date", "interval", "dateCleaned", "day")
```

```{r}
##Merge the NA averages and non NA data together
mergeData <- rbind(clean, newdata2)
```
##### 4. Make a histogram of the total number of steps taken each day 
```{r}
##Create sum of steps per date to compare with step 1
sumTable2 <- aggregate(mergeData$steps ~ mergeData$date, FUN=sum, )
colnames(sumTable2)<- c("Date", "Steps")

## Mean of Steps with NA data taken care of
stepsByDayMeanImputed <- as.integer(mean(sumTable2$Steps))

## Median of Steps with NA data taken care of
stepsByDayMedianImputed <- as.integer(median(sumTable2$Steps))
```
* Mean (Imputed): `r stepsByDayMeanImputed`
* Median (Imputed):  `r stepsByDayMedianImputed`

```{r}
## Creating the histogram of total steps per day, categorized by data set to show impact
hist(sumTable2$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Black")
hist(sumTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Grey", add=T)
legend("topright", c("Imputed Data", "Non-NA Data"), fill=c("black", "grey") )
```


## Are there differences in activity patterns between weekdays and weekends?
##### 1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
## Create new category based on the days of the week
mergeData$DayCategory <- ifelse(mergeData$day %in% c("ñóááîòà", "âîñêðåñåíüå"), "Weekend", "Weekday")

## Summarize data by interval and type of day
intervalTable2 <- ddply(mergeData, .(interval, DayCategory), summarize, Avg = mean(steps))
```

```{r}
##Plot data in a panel plot
xyplot(Avg~interval|DayCategory, data=intervalTable2, type="l",  layout = c(1,2),
       main="Average Steps per Interval Based on Type of Day", 
       ylab="Average Number of Steps", xlab="Interval")
```

